Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-10 ms/ms.tex
working on text/cover letter

As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
explicitly track the evolution of mean $\log_{10}$ SPVL
(which we denote as $\alpha$), rather than the rate of progression to
AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
to the rate of progression to AIDS; these two quantities are
deterministically linked in the model).  
In contrast to Shirreff \etal, we use a
single-stage disease model instead of accounting explicitly for
progression through the three main stages of HIV infection (primary,
asymptomatic, and disease), and we use a simple exponentially
distributed infectious period instead of a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. We account for
varying transmission rates and durations of each disease stage by
summing the durations of three stages (again based on Shirreff \etal's
model) and taking the duration-weighted average of transmission rates
of three stages. Thus the within-couple transmission rate, $\beta$,
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

@@

The uncoupled and extra-couple transmission rates (i.e., the rates of
transmission among people outside of a stable partnership, or between
people inside of a stable partnership and people other than their
partner) are scaled by
multiplying the within-couple transmission rate $\beta$ by the contact
ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
of HIV pathogenesis from three stages to a single stage could affect
our conclusions about the evolution of virulence (e.g. Kretzschmar and
Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
and multiple stages of infectivity have interactive effects on
$\rzero$). However, our simplified model produces results that are
qualitatively similar to those of Shirreff \etal's
\cite{shirreff_transmission_2011} model; when our model is calibrated
to have a similar initial epidemic growth rate $r$, the peak
\Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly
higher (4.6 \Lspvl\ vs. 4.3 \Lspvl, or 7\% higher: \figurename~\ref{fig:panel3}).

>>>>

As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
explicitly track the evolution of mean $\log_{10}$ SPVL
(which we denote as $\alpha$), rather than the rate of progression to
AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
to the rate of progression to AIDS; these two quantities are
deterministically linked in the model).  
In contrast to Shirreff \etal, we use a
single-stage disease model instead of accounting explicitly for
progression through the three main stages of HIV infection (primary,
asymptomatic, and disease), and we use an exponentially
distributed infectious period instead of a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. We account for
varying transmission rates and durations of each disease stage by
summing the durations of three stages (again based on Shirreff \etal's
model) and taking the duration-weighted average of transmission rates
of three stages. Thus the within-couple transmission rate, $\beta$,
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL
(Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

@@

The uncoupled and extra-couple transmission rates (i.e., the rates of
transmission among people outside of a stable partnership, or between
people inside of a stable partnership and people other than their
partner) are scaled by
multiplying the within-couple transmission rate $\beta$ by the contact
ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
of HIV pathogenesis from three stages to a single stage could affect
our conclusions about the evolution of virulence (e.g. Kretzschmar and
Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
and multiple stages of infectivity have interactive effects on
$\rzero$). However, our simplified model produces results that are
qualitatively similar to those of Shirreff \etal's
\cite{shirreff_transmission_2011} model; when our model is calibrated
to have a similar initial epidemic growth rate $r$, the peak
\Lspvl\ occurs at the same time ($\approx 200$ years) but slightly
higher (4.6 \Lspvl\ vs. 4.3 \Lspvl: \figurename~\ref{fig:panel3}).
